
# Latihan R di BBTNBKDS

## 2026-06-29 s/d 2026-06-30

## 1.03 Perintah dasar

## Perhitungan dasar 

Basic math:
```{r}

3 + 2
4 * 5
5 - 3

```

Spaces don't affect functionality (in most contexts).

```{r}

4/5
4 / 5
4           /              5

```

If you haven't completed a line, R will ask you to finish it.

```{r, eval = FALSE}

5 + 6 + 89 - 4 +
# finish the command on the next line, 
# or use ESC to exit

  # text or lines preceded by '#' are not executed

# lines not preceded by '#' are executed  #oops 
     
```

Use up and down arrows to move around in your history.

R displays big and small #s using normal scientific notation.

```{r}

456000000000000000
5 / 298657635

```

## Fungsi aritmatika 

```{r}

sqrt(9)
4 ^ 3
abs(-945.2368)
log(8)

# you can always use ? to get help
?log

# or get an example
example(log)
example(seq)

```

Common arithmetic functions:

 ----------------
abs(x)	                 absolute value
sqrt(x)                  square root
factorial (x)            x!
ceiling(x)	             ceiling(3.475) is 4
floor(x)	               floor(3.475) is 3
trunc(x)	               trunc(5.99) is 5
round(x, digits = n)     round(3.475, digits  2) is 3.48
signif(x, digits = n)	   signif(3.475, digits = 2) is 3.5
cos(x), sin(x), tan(x)   basic trig
log(x)	                 natural logarithm
log10(x)	               common logarithm
exp(x)	                 e^x
 ---------------


## Menetapkan variabel 

Use '<-' to assign anything you want to a variable.

```{r}

x <- 5
x

z <- "semua tak sama"
z

# mind the space!
y <- 67
y
y < - 7
y

# you can then use the variable to represent that item(s) in a formula
x * 8
y <- sqrt(4)

x * y
x == y

saya_bisa_bikin_nama_panjang <- 67

```

Important things to note regarding variables:
    case sensitive (g != G)
    should not begin with numbers or symbols (5x, %cover)
    should not contain spaces (fun_games not fun games)
    make as short and intuitive as possible

Combine 'c()' values into a list (aka vector)

```{r}

x <- c(5, 7, 8, 10)
x
# as before, the variable represents its contents in a formula
x * 2

```


## Mendapatkan bantuan

Option 1:"?function"

```{r}

?lm
?plot
# R will tell you if you look for something that doesn't exist:
?gr

```

Option 2:"help()"

```{r}

help(lm)
?help

```

option 3: "help.search("keyword")

```{r}

help.search("mle")
# the "??" shortcut gives you the same thing:
??"mle"

```

## Masukkan data 

Usually we want to read data directly in from a file. R can read in various file types. *.csv is the most commonly used one. These all do the same thing:

```{r}

my.data <- read.table(file = "1.04 temperatures.csv", 
                      header = TRUE, sep = ",")
my.data

my.data <- read.csv(file = "/Users/ajmarsha/Dropbox (Personal)/andy/teaching/BTNGP R course/1.04 temperatures.csv", header = TRUE)
my.data

my.data <- read.delim(file = "1.04 temperatures.csv", sep = ",")
my.data

my.data <- 7

```

For `read.table()`, you need to specify the separator (using "sep = "); the default separator is a space. You can drop this if you use `read.csv()`. If you use `read.delim()`, you can drop the "header = TRUE" command (this tells R that the first row is your column headings)

Common errors that thwart read commands
  - spaces in file names (e.g., 'orang.density' not 'orang density')
  - the number of variable names in row 1 != the number of columns 
          (recall != is ≠)
  - blank spaces where values are missing (use "NA")
  
```{r, eval = FALSE}

y <- read.csv("FILE_NAME",
              na.strings = c("","7","?"))  

```


## Pemeriksaan cepat ntk data yg dimasukkan 

It is good to get into the habit of frequently checking data frames you have imported or created.

```{r}

temps <- read.csv("1.04 temperatures.csv", header = TRUE)
temps

```

`dim()` gives you the dataframe's dimensions, listed as rows , columns

```{r}

dim(temps)

```

`str()` tells you the dataframe structure. ***good practice!***

```{r}

str(temps)

```

`names()` gives names of components in an object.

```{r}

names(temps)

```

`head()` and `tail()` give the first and last parts of dataframe.

```{r}

head(temps)
tail(temps)

```

Selecting subsets of data.

```{r}

#Use "$" to select subsets of data. The text before the "$" is the dataframe, the text after it is the variable name.

temps$yr
mean(temps$yr)

plot(temps$yr ~ temps$rain)

```

## Pengeluaran data 

Create a datastring: 2000 draws from a normal distribution with mean = 5, sd = 1

```{r}

my.dataset <- rnorm(n = 2000, 
                    mean = 5, sd = 1)
my.dataset

```

Use `write()` to write the data to a file. You can either provide the full path to where you want to save the file or simply name the file and it will be put in your working directory (wd).

```{r}

# specify full path
write(my.dataset, "/Users/ajmarsha/Desktop/my.dataset.csv")
write(my.dataset, "/Users/ajmarsha/Desktop/my.dataset.csv", sep = ",")
write(my.dataset, "/Users/ajmarsha/Desktop/my.dataset.txt")

# provide file name only, will be saved to wd
write(my.dataset, "my.dataset.txt")

# to check the wd
getwd()

# by default, the wd (usually) is set to the folder that holds your script file; to change, use `setwd()`. As with many things in R, it is best to give R explicit instructions rather than guess at the defaults (which might change with package updates, etc.)

setwd("/Users/ajmarsha/Desktop")
write(my.dataset, "my.dataset.txt")

```

